---
title: "Lecture 2: Literate Programming & Version Control"
subtitle: "APEC 8221: Programming for Econometrics"
author: "Ali Joglekar"
date: "September 4, 2025"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
execute:
  warning: false
  message: false
  eval: true
  echo: true
editor: visual
---

# Opening & Check-In

## Status Check

**Quick show of hands:** Who completed the Software Installation & Verification exercise successfully?

> If anyone is still having issues with Parts 1, send me an email - we'll get it sorted before Monday.

## Today's Mission

Today we're adding the final two pieces to your reproducible workflow: **Quarto** for combining code and narrative, and **Git** for version control and collaboration.

::: callout-important
You can use either **Personal GitHub** (github.com) or **UMN Enterprise GitHub** (github.umn.edu) for this course. Both work perfectly fine - it's your preference.
:::

**Git authentication is tricky** regardless of which account type you choose. Almost everyone has issues the first time. That's why we're doing this together in class rather than making you struggle alone at home.

By the end of today, you'll have your first reproducible document pushed to GitHub.

# Quarto Basics & Git Demo (`I Do`)

## Quarto Philosophy & Demo

### Quarto Motivation

-   Traditional research workflow: Write code in R, copy results to Word, manually create figures, pray nothing changes.
-   Quarto workflow: Code, narrative, and results live in one document that updates automatically.

#### Why this matters

-   No copy-paste errors between R and PowerPoint
-   Automatic updates when data changes
-   Reproducible presentations for colleagues
-   Professional workflow used in industry

#### When to use Quarto

**Dissertation & Thesis Work**

-   Reproducible chapters with embedded analysis, tables, and figures that update automatically when data changes
-   Appendices with code showing exactly how results were generated for committee review
-   Methodology sections that combine narrative explanation with actual code implementation

**Journal Article Submission**

-   Supplementary materials that journals increasingly require showing reproducible analysis
-   Replication files that many journals now mandate for publication
-   Pre-prints and working papers shared on platforms like arXiv, SSRN, or RePEc

**Conference Presentations**

-   Research presentations that seamlessly integrate live code demonstrations with slides
-   Poster sessions where QR codes link to interactive, reproducible analyses
-   Workshop materials for teaching methods to other researchers

#### Specific professional scenarios

**The "Oh No, the Data Updated" Scenario**

You're at the Minneapolis Fed. Your quarterly economic outlook report is due to the president tomorrow, but the BLS just released revised employment data.

With Quarto: change one data file, re-render, done. Without Quarto: manually update 47 charts and tables, hope you don't miss any.

**The "Explain Your Methodology" Scenario**

You're testifying before Congress about your economic impact analysis. A representative asks "How exactly did you calculate that \$2.3 billion figure?"

With Quarto: "Here's the exact code, data, and methodology - all publicly available." This builds credibility and trust.

**The "Replication Crisis" Scenario**

You're submitting to the American Economic Review. The editor asks for replication files.

With Quarto skills: send a single document that regenerates every table and figure from raw data. Without: scramble to find old code, hope it still works, spend weeks reconstructing your analysis.

**The "Team Collaboration" Scenario**

You're at a consulting firm with 5 economists working on different sections of a client report.

With Quarto: everyone works on their section, integrates seamlessly, consistent formatting. Without: nightmare of merging Word documents and rebuilding broken figures.

### Create Basic Quarto Document Demo

Quarto document consists of two parts:

**YAML**

```         
---
title: "Lecture 2 Demo: Reproducible Research"
author: "Ali Joglekar"
date: today
format: pdf
---
```

**Markdown**

````         
# Why Quarto Matters

This document combines **narrative text** with executable code.

## A Simple Analysis

```{r}
library(tidyverse)
data(mtcars)

# Calculate summary statistics
mpg_summary <- mtcars %>%
  summarise(
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    n_cars = n()
  )

print(mpg_summary)

# Store values for inline use
mean_mpg_value <- mpg_summary$mean_mpg
```

The average fuel economy is `{r} round(mean_mpg_value, 1)` miles per gallon. 

This value was calculated directly from the data and will update automatically if the data changes!


## A Simple Plot

```{r}
ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl))) +
  labs(
    title = "Horsepower vs. Fuel Economy",
    x = "Miles per Gallon",
    y = "Horsepower",
    color = "Cylinders"
  ) +
  theme_minimal()
```

This plot shows the relationship between fuel economy and engine power.
````

### Render the Document

Watch what happens when I render this... (Click Render)

See how the code runs, results appear, and everything is formatted professionally?

### What about Slides?

1.  **Change format** from `pdf` to `revealjs`
2.  **Add slide breaks** using `---`
3.  **Enhance code chunks** with:
    -   `#| echo: true` to show code
    -   `#| code-line-numbers: "1-3|4-10|12"` for progressive highlighting
    -   `#| fig-width: 8` and `#| fig-height: 5` for better plot sizing
4.  **Split content** into logical slide chunks

````         
---
title: "Lecture 2 Demo: Reproducible Research"
author: "Ali Joglekar"
date: today
format:
  revealjs:
    theme: dark
    slide-number: true
    chalkboard: true
    code-fold: false
    code-overflow: wrap
---

# Why Quarto Matters

This document combines **narrative text** with executable code.

---

## A Simple Analysis

```{r}
#| echo: true
#| code-line-numbers: "1-3|4-10|12"
library(tidyverse)
data(mtcars)

# Calculate summary statistics
mpg_summary <- mtcars %>%
  summarise(
    mean_mpg = mean(mpg),
    median_mpg = median(mpg),
    n_cars = n()
  )

print(mpg_summary)
```

```{r}
#| echo: false

# Store values for inline use
mean_mpg_value <- mpg_summary$mean_mpg
```

---

## Key Finding

The average fuel economy is `{r} round(mean_mpg_value, 1)` miles per gallon. 

This value was calculated directly from the data and will update automatically if the data changes!

---

## A Simple Plot

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5


ggplot(mtcars, aes(x = mpg, y = hp)) +
  geom_point(aes(color = factor(cyl)), size = 3) +
  labs(
    title = "Horsepower vs. Fuel Economy",
    x = "Miles per Gallon",
    y = "Horsepower",
    color = "Cylinders"
  ) +
  theme_minimal()
```

---

## Key Insight

This plot shows the relationship between fuel economy and engine power.
````

## Versioning with Git Demo

Quarto solves the 'code + narrative' problem. Git solves the 'collaboration + history' problem.

-   Git is like a sophisticated "Save As" system
-   Repository = Project folder with full history
-   Commit = Saved snapshot with description
-   Push = Upload your changes to GitHub
-   Pull = Download others' changes from GitHub

### Creating a Git Repository

#### Personal GitHub Demo:

Let me show you the basic Git workflow using Personal GitHub first - this is what most students will use.

**Create new GitHub repository on github.com:**

-   Go to github.com → New Repository
-   Name: "lecture2-demo-personal"
-   Private repository
-   Initialize with README: Yes
-   Add .gitignore: R

**Clone to RStudio:**

-   In RStudio: File → New Project → Version Control → Git

-   Repository URL: <https://github.com/%5Busername%5D/lecture2-demo-personal.git>

-   Choose location and create project

::: callout-caution
**AVOID these locations:**

-   Google Drive, OneDrive, Dropbox folders

-   Network drives (H:/ drives, etc.)

-   External USB drives

-   Any folder that syncs to the cloud

**RECOMMENDED locations:**

-   C:/Users/\[your-username\]/Documents/apec8221-\[your-x500-id\]/

-   Your local Documents folder

-   Desktop (if you don't mind it there)

**Why this matters:** Cloud sync services interfere with Git's internal files, causing corruption and mysterious errors. Since you're already using GitHub for cloud backup and collaboration, additional cloud sync is unnecessary and problematic.
:::

**Configure R for GitHub:**

```{r}
#| eval: false

# Personal GitHub is the default - no special setup needed!
library(usethis)

# Check current configuration
git_sitrep()  
# Should show "Default GitHub host: https://github.com"

# Create token for github.com
create_github_token(
  scopes = c("repo", "user", "workflow"),
  description = "Demo-Personal-Access"
)
```

**Verify it Worked:**

```{r}
#| eval: false

# Tgh::gh_whoami()

```

Should return your user information if working correctly.

#### Enterprise GitHub Demo:

Now let me show the Enterprise GitHub setup for those who prefer institutional accounts.

**Create new GitHub repository on github.umn.edu:**

-   Go to github.umn.edu → New Repository
-   Name: "lecture2-demo-enterprise"
-   Private repository
-   Initialize with README: Yes
-   Add .gitignore: R

**Clone to RStudio:**

-   In RStudio: File → New Project → Version Control → Git

-   Repository URL: <https://github.umn.edu/%5Busername%5D/lecture2-demo-enterprise.git>

-   Choose location and create project

**Configure R for Enterprise GitHub:**

```{r}
#| eval: false


# Enterprise requires special configuration
library(usethis)

# Configure .Rprofile for enterprise
usethis::edit_r_profile()

# Show adding these lines:
# options(
#   usethis.protocol = "https",
#   usethis.host = "https://github.umn.edu",
#   usethis.github_host = "https://github.umn.edu"
# )

# Restart R to load changes

# Create enterprise token
create_github_token(
  host = "https://github.umn.edu",
  scopes = c("repo", "user", "workflow"),
  description = "Demo-Enterprise-Access"
)

# Store enterprise token
gitcreds::gitcreds_set(url = "https://github.umn.edu")
```

**Verify it Worked:**

```{r}
#| eval: false


# Test your setup
gh::gh_whoami(.api_url = 'https://github.umn.edu/api/v3')
```

Should return your user information if working correctly.

### Git Cycle

This is the basic cycle: **Edit → Stage → Commit → Push**. The process is identical regardless of which GitHub account type you choose.

1.  Add file(s)
2.  Stage the file (check the box in the `Git` tab – found in Environment pane)
3.  Click **Commit** (review changes opens in a new window).
4.  Write commit message: "Add first reproducible analysis"
5.  Click **Commit**
6.  Click **Push**
7.  Revise file(s)
8.  **Stage → Commit → Push**

# Collaborative Git Setup by Account Type (`We Do`)

Quick show of hands:

-   Who is using **Personal GitHub** (github.com)? This is recommended for most students.
-   Who is using **UMN Enterprise GitHub** (github.umn.edu)?

Great! Now let's work through setup for each group.

## Personal GitHub Group:

```{r}
#| eval: false


# Personal GitHub users - this should be straightforward
library(usethis)

# Step 1: Check configuration (should already be correct)
git_sitrep()
# Should show: Default GitHub host: "https://github.com"

# Step 2: Create token (opens github.com)
create_github_token(
  scopes = c("repo", "user", "workflow"),
  description = "APEC8221-Personal-Access"
)

# Step 3: Store token
gitcreds::gitcreds_set()

# Step 4: Test connection
gh::gh_whoami()
```

## Enterprise GitHub Group:

```{r}
#| eval: false


# Enterprise GitHub users - need special configuration
library(usethis)

# Step 1: Configure .Rprofile
usethis::edit_r_profile()

# Add these lines to .Rprofile:
# options(
#   usethis.protocol = "https",
#   usethis.host = "https://github.umn.edu", 
#   usethis.github_host = "https://github.umn.edu"
# )

# Step 2: Restart R (Session → Restart R)

# Step 3: Create enterprise token
create_github_token(
  host = "https://github.umn.edu",
  scopes = c("repo", "user", "workflow"),
  description = "APEC8221-Enterprise-Access"  
)

# Step 4: Store enterprise token
gitcreds::gitcreds_set(url = "https://github.umn.edu")

# Step 5: Test enterprise connection
gh::gh_whoami(.api_url = "https://github.umn.edu/api/v3")
```

# Create Course Repository by Account Type (`You Do`)

Now you're going to create your course repository using your existing GitHub account.

::: callout-tip
Remember: **authentication issues are normal for both account types** - I'm here to help.
:::

## For Personal GitHub Users (github.com):

### Step 1: Create Repository

-   Go to **github.com**
-   Click "New Repository"
-   Name: `apec8221-[your-umn-id]` (e.g., `apec8221-smith123`)
-   Description: "APEC 8221 Programming for Econometrics coursework"
-   **IMPORTANT: Make it PRIVATE**
-   Initialize with README: Yes
-   Add .gitignore: R

### Step 2: Add Collaborator

-   Settings → Manage access → Invite collaborator
-   Username: `joglekar`
-   Send invitation

### Step 3: Clone to RStudio

-   Copy repository URL: `https://github.com/[your-username]/apec8221-[umn-id].git`
-   In RStudio: File → New Project → Version Control → Git
-   Paste URL, choose location, create project

## For Enterprise GitHub Users (github.umn.edu):

### Step 1: Create Repository

-   Go to **github.umn.edu**
-   Name: `apec8221-[your-umn-id]` (e.g., `apec8221-smith123`)
-   Description: "APEC 8221 Programming for Econometrics coursework"
-   **IMPORTANT: Make it PRIVATE**
-   Initialize with README: Yes
-   Add .gitignore: R

### Step 2: Add Collaborator

-   Username: `joglekar` (same for both account types)

### Step 3: Clone to RStudio

-   Copy repository URL: `https://github.umn.edu/[your-username]/apec8221-[umn-id].git`
-   In RStudio: File → New Project → Version Control → Git
-   Paste URL, choose location, create project

# Backup Document in GitHub (`You Do`)

## Step 1: Create First Quarto Document

-   In RStudio: File → New File → Quarto
-   Document Title: "APEC 8221 - My Research Workflow"
-   Save as: hello_quarto.qmd

**Use this content:**

````         
---
title: "APEC 8221 - My Research Workflow"
author: "[Your Name]"
date: today
format: pdf
---

# Introduction

This is my first reproducible document for APEC 8221 using [Personal/Enterprise] GitHub.

## Basic R Skills

```{r}
# Demonstrate basic R commands
library(tidyverse)

# Create some data
my_data <- data.frame(
  x = 1:10,
  y = (1:10)^2 + rnorm(10, 0, 5)
)

# Summary statistics
summary(my_data)

# Store values for inline use
mean_x <- mean(my_data$x)
```

The average of variable `x` is `{r} round(mean_x, 1)`. 

## My First Plot

```{r}
ggplot(my_data, aes(x = x, y = y)) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "My First Reproducible Plot",
    x = "X Variable",
    y = "Y Variable"
  ) +
  theme_minimal()
```
````

## Step 2: Git Workflow Practice (Same for Both Account Types)

1.  Render the document (make sure it works!)
2.  Look at Git tab - see the new files?
3.  Stage the files (check boxes for `.qmd` and `.pdf/.html`)
4.  Commit with message: "Add first reproducible analysis with \[Personal/Enterprise\] GitHub"
5.  Push to GitHub (authentication happens here!)

## Account-Specific Troubleshooting During Exercise:

### Personal GitHub Issues:

-   **Token problems:** Make sure `create_github_token()` opened [github.com](https:/github.com) (not [github.umn.edu](https://github.umn.edu))
-   **Repository URL:** Should start with `https://github.com/` (not github.umn.edu)
-   **Test command:** `gh::gh_whoami()` (no additional parameters needed)

### Enterprise GitHub Issues:

-   **Wrong host:** Check that `git_sitrep()` shows github.umn.edu as default host
-   **Token problems:** Make sure `create_github_token(host = "https://github.umn.edu")` opened the enterprise site
-   **Repository URL:** Should start with `https://github.umn.edu/`
-   **Test command:** `gh::gh_whoami(.api_url = "https://github.umn.edu/api/v3")`

# Wrap-up

## Check Success by Account Type:

Quick show of hands:

-   Personal GitHub users who successfully pushed to github.com?
-   Enterprise GitHub users who successfully pushed to github.umn.edu?

If you didn't get there today with either account type, **don't panic**. Post your issue on the Canvas discussion board or email me your exact error message **and which account type you're using**. We'll get it working.

## Assignment 0 Part 4:

Now you can complete the rest of Assignment 0.

## Reading Assignment for Tuesday:

Tuesday we dive into data manipulation with `dplyr` - the heart of `tidyverse`.

For Tuesday's class on data manipulation, please read:

-   **R4DS Chapter 3 (Data Transformation)** - the core `dplyr` verbs we'll use constantly

-   **R4DS Chapter 7 (Data Import)** - reading files with `readr`

-   **R4DS Chapter 27.2-27.3 (A Field Guide to Base R)** - selecting elements from vectors and data frames

Chapter 3 is the most important - focus there if you're short on time. We'll be using `filter()`, `select()`, `mutate()`, `group_by()`, and `summarise()` extensively.
