---
title: "APEC 8221 - Assignment 2: Automation and API Data Acquisition"
subtitle: "Submission File"
author: "[Runcheng Xu]"
date: "`r Sys.Date()`"
format: 
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    code-overflow: wrap
execute:
  warning: false
  message: false
  echo: true
  eval: true
editor: visual
---

**GitHub Repository URL:** <https://github.com/Runcheng112/apec8221-xu001296.git>

# Build Your API Function

```{r}
#| label: setup
#| message: false

# Load required packages
library(tidyverse)
library(httr2)
library(jsonlite)



# Secure API key storage – load from .Renviron file
fred_api_key <- Sys.getenv("FRED_API_KEY")


```

## API Setup and Testing

```{r}
#| label: api-testing

# Test basic API connectivity
# Your code here to test the FRED API connection
base_url <- "https://api.stlouisfed.org/fred/series/observations"
test_url <- paste0(base_url, "?", "series_id=UNRATE",
"&api_key=", fred_api_key, "&file_type=json")
# Make the request
response <- request(test_url) %>% req_perform()
# Check if it worked
resp_status(response) # Should be 200
# Look at the JSON structure
raw_data <- response %>% resp_body_json()
str(raw_data, max.level = 2)

```

## Write Your Data Acquisition Function

```{r}
#| label: main-function


library(httr2)
library(dplyr)

get_fred_unemployment <- function(state_code, api_key) {
  # Input validation
  if (!grepl("^[A-Z]{2}$", state_code)) {
    stop("state_code must be two capital letters, e.g., 'CA' or 'NY'")
  }
  
  series_id <- paste0(state_code, "URN")
  base_url <- "https://api.stlouisfed.org/fred/series/observations"
  request_url <- paste0(
    base_url,
    "?series_id=", series_id,
    "&api_key=", api_key,
    "&file_type=json"
  )
  
  response <- request(request_url) %>% req_perform()
  
  if (resp_status(response) != 200) {
    stop(paste("API request failed. Status:", resp_status(response)))
  }
  
  raw_data <- response %>% resp_body_json()
  if (is.null(raw_data$observations)) {
    stop("No data returned. Check state_code.")
  }
  
  result <- raw_data$observations %>%
    purrr::map_df(~{
      data.frame(
        date = as.Date(.x$date),
        unemployment_rate = as.numeric(.x$value),
        state = state_code,
        stringsAsFactors = FALSE
      )
    })
  
  return(result)
}

# Test
test_result <- get_fred_unemployment("CA", fred_api_key)
head(test_result)



```

# Automate Data Collection

## Define Your State List

```{r}
#| label: state-selection

# Select 8-10 states from different regions
target_states <- c("CA", "TX", "NY", "FL", "OH", "WA", "CO", "GA")

# Print your selected states
```

## Implement Rate Limiting

```{r}
#| label: rate-limiting-function

library(httr2)
library(dplyr)

get_fred_unemployment <- function(state_code, api_key) {
  # Input validation
  if (!grepl("^[A-Z]{2}$", state_code)) {
    stop("state_code must be two capital letters, e.g., 'CA' or 'NY'")
  }
  
  cat("Downloading data for", state_code, "...\n")
  
  Sys.sleep(1)  # wait 1 second before API call
  
  series_id <- paste0(state_code, "URN")
  base_url <- "https://api.stlouisfed.org/fred/series/observations"
  request_url <- paste0(
    base_url,
    "?series_id=", series_id,
    "&api_key=", api_key,
    "&file_type=json"
  )
  
  response <- request(request_url) %>% req_perform()
  
  # Check response status
  if (resp_status(response) != 200) {
    stop(paste("API request failed. Status:", resp_status(response)))
  }
  
  raw_data <- response %>% resp_body_json()
  if (is.null(raw_data$observations)) {
    stop("No data returned. Check state_code.")
  }
  
  result <- raw_data$observations %>%
    purrr::map_df(~{
      data.frame(
        date = as.Date(.x$date),
        unemployment_rate = as.numeric(.x$value),
        state = state_code,
        stringsAsFactors = FALSE
      )
    })
  
  return(result)
}

```

## Compare Iteration Methods

### Method 1: `for` Loop Approach

```{r}
#| label: method-for-loop

# Initialize empty data frame
unemployment_data_loop <- data.frame()

# Loop through states
for (state in target_states) {
  cat("Downloading data for", state, "...\n")
  
  # Get data for this state
  state_data <- get_fred_unemployment(state, fred_api_key)
  
  # Add it to unemployment_data_loop
  unemployment_data_loop <- rbind(unemployment_data_loop, state_data)
  
  # Pause to respect API rate limits
  Sys.sleep(1)
}

# Check results
head(unemployment_data_loop)

# Check results
```

### Method 2: `lapply` Approach

```{r}
#| label: method-apply

unemployment_list <- lapply(target_states, function(state) {
  cat("Downloading data for", state, "...\n")
  Sys.sleep(1)  # pause to respect API limits
  get_fred_unemployment(state, fred_api_key)
})

unemployment_data_apply <- do.call(rbind, unemployment_list)

head(unemployment_data_apply)

```

### Method 3: purrr::map Approach

```{r}
#| label: method-purrr
library(purrr)
library(dplyr)

# Use purrr::map to get data for all states
unemployment_data_purrr <- target_states %>%
  map(function(state) {
    cat("Downloading data for", state, "...\n")
    Sys.sleep(1)  # pause to respect API limits
    get_fred_unemployment(state, fred_api_key)
  }) %>%
  bind_rows()

# Check results
head(unemployment_data_purrr)

```

### Method 4: Manual Approach (Demonstration)

```{r}
#| label: method-manual
#| eval: false


# Get data for each state one by one
ca_data <- get_fred_unemployment("CA", fred_api_key)
tx_data <- get_fred_unemployment("TX", fred_api_key)
ny_data <- get_fred_unemployment("NY", fred_api_key)
fl_data <- get_fred_unemployment("FL", fred_api_key)
oh_data <- get_fred_unemployment("OH", fred_api_key)


unemployment_data_manual <- rbind(ca_data, tx_data, ny_data, fl_data, oh_data)

# Check results
head(unemployment_data_manual)

```

**Iteration Method Comparison:** The `for` loop is simple and explicit, but less concise. `lapply` is more efficient and compact, while `purrr::map` is the most readable and integrates well with tidyverse workflows. Automation is valuable because it reduces errors and saves time when working with multiple states.\

**Scaling Analysis:**\
For all 50 states, the same code can be used by extending the state list. The main challenges are API rate limits and longer run times. `for` loops may become slow, while `lapply` and `purrr::map` handle larger scale more efficiently.

------------------------------------------------------------------------

# Data Analysis and Visualization

```{r}
#| label: prepare-data

# Use the data from your preferred method above
unemployment_data <- unemployment_data_loop  # or unemployment_data_apply or ...

# Check data structure
glimpse(unemployment_data)
```

## Regional Unemployment Patterns

### Basic Summary Statistics

```{r}
#| label: summary-statistics


library(dplyr)

state_averages <- unemployment_data %>%
  filter(date >= "2019-01-01" & date <= "2024-12-31") %>%
  group_by(state) %>%
  summarise(avg_rate = mean(unemployment_rate, na.rm = TRUE)) %>%
  arrange(desc(avg_rate))

highest_peak <- unemployment_data %>%
  filter(date >= "2019-01-01" & date <= "2024-12-31") %>%
  group_by(state) %>%
  summarise(max_rate = max(unemployment_rate, na.rm = TRUE)) %>%
  arrange(desc(max_rate)) %>%
  slice(1)

volatility_analysis <- unemployment_data %>%
  filter(date >= "2019-01-01" & date <= "2024-12-31") %>%
  group_by(state) %>%
  summarise(sd_rate = sd(unemployment_rate, na.rm = TRUE)) %>%
  arrange(desc(sd_rate)) %>%
  slice(1)

print(state_averages)
print(highest_peak)
print(volatility_analysis)


```

**Key Findings:**

**CA** has the highest average unemployment rate (≈5.98%), while **GA** has the lowest (≈4.02%)

**NY** shows both the highest peak (16.7%) and the greatest volatility (SD ≈2.82).

## Time Series Visualization

```{r}
#| label: time-series-plot
#| fig-width: 10
#| fig-height: 6

# Create time series plot
unemployment_plot <- unemployment_data %>%
  filter(date >= as.Date("2019-01-01")) %>%  
  ggplot(aes(x = date, y = unemployment_rate, color = state)) +
  geom_line(size = 1) +  
  geom_vline(xintercept = as.Date("2020-03-01"), 
             linetype = "dashed", color = "red") +   
  labs(title = "Unemployment Trends by State (2019–2024)",
       subtitle = "Dashed red line marks COVID start (March 2020)",
       x = "Date",
       y = "Unemployment Rate (%)",
       color = "State") +
  theme_minimal(base_size = 14) +  
  theme(legend.position = "bottom")

print(unemployment_plot)
```

**Analysis:** Unemployment rates spiked sharply across all states in March 2020, reflecting the immediate impact of COVID-19. Rates then gradually declined, though the speed of recovery varied by state. By 2022–2024, unemployment stabilized at lower levels, but small differences across states remained.

## State Comparison Analysis

```{r}
#| label: comparison-analysis
library(dplyr)
library(ggplot2)

unemployment_comparison <- unemployment_data %>%
  mutate(period = case_when(
    date >= as.Date("2019-01-01") & date <= as.Date("2020-02-29") ~ "Pre-COVID",
    date >= as.Date("2023-01-01") & date <= as.Date("2024-12-31") ~ "Post-COVID",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(state, period) %>%
  summarise(avg_rate = mean(unemployment_rate, na.rm = TRUE), .groups = "drop")

unemployment_comparison <- unemployment_comparison %>%
  group_by(state) %>%
  arrange(state, desc(period)) %>%
  ungroup() %>%
  mutate(state = reorder(state, avg_rate))

comparison_plot <- ggplot(unemployment_comparison, aes(x = state, y = avg_rate, fill = period)) +
  geom_col(position = "dodge") +
  labs(title = "Average Unemployment Rate by State (Pre vs Post COVID)",
       x = "State",
       y = "Average Unemployment Rate (%)",
       fill = "Period") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(comparison_plot)


```

**Analysis:**\

The bar chart shows that unemployment rates generally increased in the Post-COVID period (2023–2024) compared to the Pre-COVID period (2019–early 2020). States like California (CA) and Colorado (CO) saw the largest increases, while Florida (FL) remained relatively stable. This suggests that COVID had long-term effects on unemployment, with stronger impacts in some states than others.

------------------------------------------------------------------------

# Extension and Critical Thinking

## API Pagination Challenge

### Step 1: Explore API Metadata

```{r}
#| label: pagination-exploration

# Explore pagination with a limited request
response <- request("https://api.stlouisfed.org/fred/series/observations") %>%
  req_url_query(
    series_id = "CAUR",   # California unemployment
    api_key = fred_api_key,
    file_type = "json",
    limit = 100           # Small limit to see pagination
  ) %>%
  req_perform()

full_data <- response %>% resp_body_json()

# Examine metadata structure
str(full_data, max.level = 1)

```

**Metadata Analysis:** **Metadata Analysis**

The FRED API provides pagination information through the parameters **`limit`** (how many observations per call) and **`offset`** (starting index for the data). The metaata also shows **`count`**, which indicates the **total number of available observations** (here it is 596). In the first call, with `limit = 100`, the API returned **100 observations** out of the total 596.

### Step 2: Build Complete Data Function

```{r}
#| label: pagination-function
get_fred_unemployment_complete <- function(state_code, api_key, page_size = 100) {
  
  # Step 1: Get metadata to find total number of observations
  meta_response <- request("https://api.stlouisfed.org/fred/series/observations") %>%
    req_url_query(
      series_id = state_code,
      api_key = api_key,
      file_type = "json",
      limit = 1   # only need 1 record to get metadata
    ) %>%
    req_perform()
  
  meta_data <- meta_response %>% resp_body_json()
  total_obs <- as.numeric(meta_data$count)
  
  message("Total observations available: ", total_obs)
  
  # Step 2: Calculate number of pages needed
  num_pages <- ceiling(total_obs / page_size)
  message("Fetching data in ", num_pages, " pages...")
  
  # Step 3: Loop through pages and collect data
  all_data <- list()
  
  for (i in 0:(num_pages - 1)) {
    offset_val <- i * page_size
    
    message("Getting page ", i + 1, " of ", num_pages, " (offset = ", offset_val, ")")
    
    page_response <- request("https://api.stlouisfed.org/fred/series/observations") %>%
      req_url_query(
        series_id = state_code,
        api_key = api_key,
        file_type = "json",
        limit = page_size,
        offset = offset_val
      ) %>%
      req_perform()
    
    page_data <- page_response %>% resp_body_json()
    
    # Extract observations
    all_data[[i + 1]] <- tibble::tibble(
      date = as.Date(sapply(page_data$observations, function(x) x$date)),
      value = as.numeric(sapply(page_data$observations, function(x) x$value))
    )
  }
  
  # Step 4: Combine and return complete dataset
  complete_data <- dplyr::bind_rows(all_data)
  return(complete_data)
}

ca_data <- get_fred_unemployment_complete("CAUR", fred_api_key, page_size = 100)
head(ca_data)
```

### Step 3: Pagination Analysis

```{r}
#| label: pagination-analysis

# Compare complete vs. recent data
# Your analysis code here

# Answer the required questions:
# 1. How many total observations for California?
# 2. How many API calls needed with page_size = 500?  
# 3. Trade-offs between page sizes?
# 4. When to use complete vs. recent data?
```

**Pagination Insights:**

1.  For California unemployment data, there are **596 total observations** available, as shown in the API metadata
2.  If a **page size of 500** is used, the dataset would require **2 API calls**: the first call would retrieve 500 records, and the second call would fetch the remaining 96 records.
3.  The trade-offs between page sizes involve efficiency versus reliability. **Larger page sizes** reduce the number of API calls, improving performance, but they may risk timeouts or memory issues if the dataset is very large. **Smaller page sizes**, on the other hand, are more reliable and easier to handle, but they require more calls, which can slow down processing and increase server load.
4.  **Complete historical data** is useful for long-term trend analysis, policy evaluation, and studying structural changes in the labor market. **Recent data only** is more appropriate for short-term forecasting, real-time monitoring, or analyzing current economic conditions without the overhead of handling the entire historical dataset.

## Alternative API Practice

**Selected API:** *Option B (BLS）*

### API Documentation Research

The BLS API requires sending series IDs and an API key in the request body, and the response is returned in a nested JSON format with results grouped under `"Results"` and `"series"`. This structure is different from FRED, which uses query parameters in the URL and provides metadata like observation counts and pagination controls more explicitly.

### Custom API Function

```{r}
#| label: alternative-api-function
bls_api_key <- "74c72444bdac46fd9977d3af407c9d7e"
library(httr)
library(jsonlite)
library(dplyr)

bls_api_key <- "74c72444bdac46fd9977d3af407c9d7e"

# Custom function for BLS API
get_bls_data <- function(series_id, startyear = 2019, endyear = 2024, api_key = bls_api_key) {
  url <- "https://api.bls.gov/publicAPI/v2/timeseries/data/"
  
  response <- POST(
    url,
    body = list(
      seriesid = list(series_id),
      startyear = startyear,
      endyear = endyear,
      registrationkey = api_key
    ),
    encode = "json"
  )
  
  data <- content(response, as = "parsed", type = "application/json")
  
  df <- data.frame(
    year = sapply(data$Results$series[[1]]$data, function(x) x$year),
    period = sapply(data$Results$series[[1]]$data, function(x) x$period),
    value = sapply(data$Results$series[[1]]$data, function(x) x$value)
  )
  
  df <- df %>%
    filter(grepl("M", period)) %>% # 
    mutate(
      date = as.Date(paste0(year, "-", substring(period, 2), "-01")),
      value = as.numeric(value)
    ) %>%
    arrange(date)
  
  return(df)
}

# 🔹 Test your function
test_result <- get_bls_data("CES0500000003", startyear = 2019, endyear = 2024)
head(test_result)

```

### Data Collection and Analysis

```{r}
#| label: alternative-api-collection

library(dplyr)
library(purrr)

target_entities <- c("CES0500000003",  # Total private hourly earnings
                     "CES0600000003",  # Manufacturing hourly earnings
                     "CES0800000003")  # Construction hourly earnings

api_data <- map_df(target_entities, function(id) {
  df <- get_bls_data(id, startyear = 2019, endyear = 2024)
  df$series_id <- id
  return(df)
})

summary_stats <- api_data %>%
  group_by(series_id) %>%
  summarise(
    avg_value = mean(value, na.rm = TRUE),
    max_value = max(value, na.rm = TRUE),
    min_value = min(value, na.rm = TRUE)
  )

head(summary_stats)

```

### Visualization

```{r}
#| label: alternative-api-viz
#| fig-width: 10
#| fig-height: 6

library(ggplot2)

api_plot <- ggplot(api_data, aes(x = date, y = value, color = series_id)) +
  geom_line(size = 1) +
  labs(
    title = "BLS Employment/Wage Trends",
    x = "Date",
    y = "Value",
    color = "Series ID"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

print(api_plot)

```

### API Comparison and Reflection

Working with the BLS API was more challenging than FRED because the data comes in a nested JSON structure and requires handling request bodies rather than just URL queries. This taught me that different APIs vary in format and complexity, so flexibility and careful reading of documentation are key when adapting code.

## API Landscape Survey

### Quick API Exploration

```{r}
#| label: api-survey

# Explore three different API structures without writing full functions

# API 1: NASA Astronomy Picture of the Day (nested JSON, government API)
nasa_response <- request("https://api.nasa.gov/planetary/apod?api_key=DEMO_KEY") %>% 
  req_perform()
nasa_data <- nasa_response %>% resp_body_json()
str(nasa_data)

# API 2: JSONPlaceholder (flat JSON, RESTful design)
posts_response <- request("https://jsonplaceholder.typicode.com/posts/1") %>% 
  req_perform()
posts_data <- posts_response %>% resp_body_json()
str(posts_data)

# API 3: Cat Facts (simple response, no authentication)
facts_response <- request("https://catfact.ninja/fact") %>% 
  req_perform()
facts_data <- facts_response %>% resp_body_json()
str(facts_data)
```

### API Structure Analysis

1.   The response structures differ from FRED because FRED returns tabular time-series data, while NASA uses nested JSON with text and URLs, JSONPlaceholder gives flat JSON, and Cat Facts provides a very simple one-field JSON.

2.   NASA requires an API key (though a demo key works), while JSONPlaceholder and Cat Facts require no authentication

3.   Government APIs (like NASA, FRED, BLS) are more structured and secure, commercial APIs often include pagination and strict key use, while fun APIs are lightweight and mainly for practice.

4.   Across APIs, common patterns include JSON responses, frequent use of keys for authentication, and parameters for pagination or limits, but structures range from highly nested to very flat.
