---
title: "Lecture 5: Automation – Writing Functions"
subtitle: "APEC 8221: Programming for Econometrics"
author: "Ali Joglekar"
date: "September 16, 2025"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  eval: true
  echo: true
editor: visual
---

# Today's Mission

Today we're focusing on **writing functions** to eliminate repetitive code. You'll learn the DRY principle and basic iteration techniques that make your research more efficient and professional.

**Why functions matter:** If you're copy-pasting code with minor modifications, you're doing it wrong. Functions eliminate repetition, reduce errors, and make your code reusable.

# The DRY Principle & Function Basics (`I Do`)

## What is DRY?

**DRY = "Don't Repeat Yourself"**

Let me show you with a real example. You're analyzing unemployment by region:

```{r}
#| label: setup-data

library(tidyverse)

# Simulate unemployment data
set.seed(123)
unemployment_data <- data.frame(
  year = rep(2000:2020, 4),
  region = rep(c("Northeast", "Southeast", "Midwest", "West"), each = 21),
  unemployment_rate = c(
    rnorm(21, 5.2, 1.2),  # Northeast
    rnorm(21, 6.1, 1.5),  # Southeast  
    rnorm(21, 5.8, 1.4),  # Midwest
    rnorm(21, 5.5, 1.3)   # West
  )
)
```

::: callout-tip
## `set.seed()`

-   Set a seed to ensure reproducibility when using functions that involve random number generation (e.g., `rnorm()`, `runif()`, `sample()`).

-   With the same seed, these functions will produce the exact same sequence of "random" numbers each time the code is executed

-   Any integer can be used as the seed value

```{r}
#| label: set-seed

# Set the seed to 123
set.seed(123)

# Generate a sequence of 5 random normal numbers
random_numbers_1 <- rnorm(5)
print(random_numbers_1)

# Set the same seed again
set.seed(123)

# Generate another sequence of 5 random normal numbers
random_numbers_2 <- rnorm(5)
print(random_numbers_2)

# Set a different seed
set.seed(1234)

# Generate a sequence of 5 random normal numbers
random_numbers_3 <- rnorm(5)
print(random_numbers_3)
```
:::

### The Bad Way (Copy-Paste)

```{r}
#| label: bad-way

# Northeast analysis
northeast_data <- unemployment_data %>% 
  filter(region == "Northeast")
ne_mean <- mean(northeast_data$unemployment_rate)
ne_sd <- sd(northeast_data$unemployment_rate)
paste("Northeast: Mean =", round(ne_mean, 2), "SD =", round(ne_sd, 2), sep = " ")

# Southeast analysis  
southeast_data <- unemployment_data %>% 
  filter(region == "Southeast")
se_mean <- mean(southeast_data$unemployment_rate)
se_sd <- sd(southeast_data$unemployment_rate)
paste("Southeast: Mean =", round(se_mean, 2), "SD =", round(se_sd, 2), sep = " ")

# This is getting tedious...
```

### The Good Way (Function)

```{r}
#| label: good-way

# Write once, use everywhere
analyze_region <- function(data, region_name) {
  result <- data %>% 
    filter(region == region_name) %>%
    summarise(
      mean = mean(unemployment_rate),
      sd = sd(unemployment_rate)
    )
  
  return(result)
}

# Use for all regions
regions <- c("Northeast", "Southeast", "Midwest", "West")
for (region in regions) {
  result <- analyze_region(unemployment_data, region)
  cat(region, ": Mean = ", round(result$mean, 2), 
      ", SD = ", round(result$sd, 2), "\n", sep="")
}
```

::: callout-tip
## `print()` versus `cat()`: How They Display Text

**`print()`** shows R objects with extra formatting:

```{r}
#| label: print-text

print("Hello world") 
print(c("apple", "banana"))   
```

**`cat()`** displays raw text, no extra formatting:

```{r}
#| label: cat-text

cat("Hello world\n") 
cat("apple", "banana") 
```

-   `print()` is for showing R objects (with formatting)
-   `cat()` is for displaying clean text output – great for presenting results from functions
    -   `\n` represents the newline character
:::

## Function Anatomy

An R function is an object that does something for you.

Parts of a function:

1.  **Name:** they are R objects

2.  **Arguments:** the inputs, which are other R objects

3.  **Body:** the code that does the work

4.  **Return:** the planned result of all that work

    ➔ Will automatically return the last evaluated expression, but it's still a good habit to use `return()` for explicitness

```{r}
#| label: function-anatomy
#| eval: false

function_name <- function(argument1, argument2 = default) {
  # Do the work
  result <- some_calculation(argument1, argument2)
  return(result)
}
```

::: callout-important
**When to write functions:** If you copy-paste code 3+ times, write a function instead.
:::

## How R Gobbles Your Code

1.  **Line breaks:**

    When R is reading code and gets to the end of the line,

    -   It does what the line says if it has accumulated a correct and complete statement.

    -   Otherwise it keeps reading on the next line. To continue a statement on the next line,

        -   End with a comma, `+`, `-`, `*`, `/`, etc.
        -   Or leave parentheses unclosed

2.  **Expression blocks (compound expressions):**

    -   When R sees a left brace, `{`, it gobbles up the code, but does not start evaluation until it sees a matching right brace, `}`.
    -   Then it evaluates everything and returns whatever the last line produces.

# Live Demo: Basic Iteration (`I Do`)

Let's build a function and apply it using different iteration approaches:

```{r}
#| label: demo-data

# Economic growth data
set.seed(456)
countries <- c("Germany", "France", "Italy", "Spain")
years <- 2015:2020

gdp_data <- expand.grid(country = countries, year = years) %>%
  arrange(country, year) %>%
  group_by(country) %>%
  mutate(
    gdp_per_capita = 25000 * (1.02^(year - 2015)) * (1 + rnorm(n(), 0, 0.02))
  ) %>%
  ungroup()

head(gdp_data, 8)
```

## Write the Function

```{r}
#| label: demo-function

calculate_growth_stats <- function(data, country_name) {
  country_data <- data %>%
    filter(country == country_name) %>%
    arrange(year) %>%
    mutate(growth_rate = (gdp_per_capita / lag(gdp_per_capita) - 1) * 100)
  
  avg_growth <- mean(country_data$growth_rate, na.rm = TRUE)
  
  return(avg_growth)
}

# Test it
germany_result <- calculate_growth_stats(gdp_data, "Germany")
cat("Germany average growth:", round(germany_result, 2), "%\n")

italy_result <- calculate_growth_stats(gdp_data, "Italy")
cat("Italy average growth:", round(italy_result, 2), "%\n")
```

## What Happens in a Function Stays in a Function

Without taking special steps, any object created inside the function disappears when the function concludes.

```{r}
#| label: inside-fxn
#| error: true

# Try to print the avg_growth object from our calculate_growth_stats function
avg_growth

# What's in the global environment? 
# No `avg_growth` 
ls()
```

If you say `Z <- f(Z)`, the value of `Z` is not changed by `f()`, but by `<-`. Changing `Z` inside the function has no effect on `Z` outside the function.

## Three Ways to Apply It

### Method 1: `for` Loop

`countries_list` is a vector and `cntr` takes on each of its value during the loop. In each iteration, the body of the function is evaluated and the result is stored in `growth_summary`

```{r}
#| label: demo-for-loop

# extract vector of country names (could have also just used `countries`)
countries_list <- unique(gdp_data$country)

# create shell for data results
growth_summary <- data.frame(country = character(), avg_growth = numeric())

# 
# will evaluate the body of the loop iteratively over each item 
#   in `countries_list`using cntr = countries_list[1], 
#   then cntr = countries_list[2], ..., cntr = countries_list[n]
#   where n = length(countries_list) --> 4
#
#     countries_list[1] == "Germany"
#     countries_list[2] == "France"
#     countries_list[3] == "Italy"
#     countries_list[4] == "Spain"

for (cntr in countries_list) {
  # run our calculate_growth_stats function where data == gdp_data and country == cntr
  # remember result object is a list
  result <- calculate_growth_stats(gdp_data, cntr)
  
  # save results in 1x2 data.frame with country name
  result_df <- data.frame(
    country = cntr, 
    avg_growth = result
  )
  
  # row bind (`rbind`) result_df to growth_summary object
  # growth_summary object will continue to build up over each iteration
  growth_summary <- rbind(growth_summary, result_df)
}

print(growth_summary)
```

### Method 2: `lapply`/`sapply`

`lapply(X, FUN, ...)`

1.  `X` is a list, `FUN` is a function.
2.  Translation: apply `FUN` to each element of `X`.
    -   `lapply(DF, sd)` → list of the standard deviation of each variable in `DF`.
    -   In each iteration R does `sd(DF[[j]])` (`j` is what changes.)
3.  `lapply` always returns a list.
4.  **Problem:** What if some variables have `NA` values? The `...` argument is there to let you provide any extra arguments that `FUN` needs.
    -   `lapply(DF, sd, na.rm=TRUE)` → In each iteration

```{r}
#| label: demo-lapply

# lapply returns list of single values
growth_list <- lapply(countries_list, function(country) {
  calculate_growth_stats(gdp_data, country)
})
growth_list

unlist(growth_list) # returns a named vector
```

`sapply(X, FUN, ...)`

1.  `lapply` always returns a list, but often a vector would be more convenient.
2.  `sapply` converts the list to a more convenient object (if it makes sense).
    -   If each item in the list has length 1 and all are the same type, `sapply` returns a vector. Example: `sapply(DF, sd, na.rm=TRUE)`
    -   Otherwise things can be more complicated, so CHECK.

```{r}
#| label: demo-sapply

# sapply extracts and simplifies
growth_rate <- sapply(countries_list, function(country) {
  calculate_growth_stats(gdp_data, country)
})
growth_rate
```

If we want to get a format similar to the output of our `for` loop above:

```{r}
#| demo-format

# Convert to data frame
growth_summary <- data.frame(
  country = countries_list,
  growth_rate = growth_rate
)

print(growth_summary)
```

::: callout-tip
## Anonymous (lambda) Functions

These are functions that are defined, but not assigned a name

They are usually used with the `*apply` functions.

**Example:** Count the number of NAs in every variable of a data frame

```{r}
#| label: lamda-fxn
#| eval: false

sapply(DF, function(x){sum(is.na(x))})
```

A recent version of R introduced `\(x)` as a synonym for `function(x)`.

```{r}
#| label: lamda-fxn2
#| eval: false

sapply(DF, \(x){sum(is.na(x))})
```

1.  R passes the anonymous function one column at a time
2.  The function returns a count of NAs
3.  `sapply` assembles those counts into nice little vector
4.  The function disappears
:::

### Method 3: `apply` (for matrices)

```{r}
#| label: demo-matrix-apply

# Create a simple matrix from our GDP data
# Just use the first few countries and years for simplicity
cntrs <- c("Germany", "France", "Italy")
yrs <- gdp_data %>% 
  filter(year >=2015) %>%
  pull(year) %>%
  unique()

gdp_subset <- gdp_data %>%
  filter(country %in% cntrs, year %in% yrs) %>%
  arrange(country, year) %>%  # Make sure data is ordered properly
  select(gdp_per_capita) %>%
  pull() %>%  # Extract the vector
  matrix(nrow = 6, ncol = 3)  # 6 years, 3 countries
gdp_subset

# Add column & row names for clarity
colnames(gdp_subset) <- cntrs
rownames(gdp_subset) <- yrs
gdp_subset

# apply() works on matrix margins: 1 = rows
year_means <- apply(gdp_subset, 1, mean, na.rm = TRUE)  # Row means
print(round(year_means, 0))

# apply() works on matrix margins: 2 = columns
country_means <- apply(gdp_subset, 2, mean, na.rm = TRUE)  # Column means
print(round(country_means, 0))
```

## Quick Comparison

| Method | Best for | When to use |
|----|----|----|
| **`for` loops** | Learning, debugging, complex logic | Always a good choice |
| **`lapply`/`sapply`** | Lists/vectors, simple operations | Base R workflows |
| **`apply`** | Matrices/arrays | Econometric results, correlation matrices |

# Debugging Interlude: Function Problems

Let's see the most common function mistakes:

## Error 1: Syntax errors (unmatched parentheses / braces)

```{r}
#| label: debugging-functions1
#| error: true

bad_function <- function(x) {
  result <- x * 2
  return(result)

good_function <- function(x) {
  result <- x * 2
  return(result)
}
```

## Error 2: Wrong number of arguments

```{r}
#| label: debugging-functions2
#| error: true

calculate_growth <- function(start_value, end_value, years) {
  ((end_value / start_value)^(1/years) - 1) * 100
}

calculate_growth(100, 150)  # Missing 3rd 'years' argument
```

## Error 3: Variable scope - objects not found outside function

```{r}
#| label: debugging-functions3
#| error: true

# Create simple test data
my_data <- data.frame(
  value = c(10, 15, NA, 20, 25, NA, 30)
)

process_data <- function(data) {
  cleaned_data <- data %>% filter(!is.na(value))
  summary_stats <- mean(cleaned_data$value)
  return(summary_stats)
}

result <- process_data(my_data)

# Cannont access variables created inside the function
print(cleaned_data)
print(summary_stats)
```

**Your debugging checklist**

1.  **COUNT** parentheses and braces - they must match
2.  **CHECK** function arguments - provide all required parameters
3.  **REMEMBER** scope - variables inside functions aren't available outside
4.  **TEST** with simple inputs first

# Guided Practice: Building Together (`We Do`)

Let's build a function to calculate different types of growth rates:

```{r}
#| label: we-do-setup

# Sample data
economic_data <- data.frame(
  year = 2015:2020,
  gdp = c(20000, 20500, 21200, 21800, 22100, 22900),
  employment = c(148, 151, 153, 156, 158, 160)
)

head(economic_data)
```

## Interpret Growth Function Together

This function calculates either simple percentage change or compound annual growth rate, with built-in error checking for invalid inputs.

Let's unpack the function line-by-line. What should this function do?

```{r}
#| label: we-do-function
#| error: true

calculate_growth <- function(start_value, end_value, years = NULL, method = "simple") {
  
  if (method == "simple") {
    growth_rate <- (end_value - start_value) / start_value * 100
  } else if (method == "cagr") {
    if (is.null(years)) stop("Need years for CAGR")
    growth_rate <- ((end_value / start_value)^(1/years) - 1) * 100
  } else {
    stop("Method must be 'simple' or 'cagr'")
  }
  
  return(growth_rate)
}

# Test it
#-------------

# simple growth rate
gdp_simple <- calculate_growth(20000, 22900)   
cat("GDP simple growth:", round(gdp_simple, 1), "%\n")

# compound annual growth rate
gdp_cagr <- calculate_growth(20000, 22900, years = 5, method = "cagr")
cat("GDP CAGR:", round(gdp_cagr, 1), "%\n")

# average annual growth rate
gdp_aagr <- calculate_growth(20000, 22900, years = 5, method = "aagr")  
```

::: {.callout-note collapse="true"}
## `calculate_growth()` Function Interpretation

### **Function definition:**

`calculate_growth <- function(...)`

Creates a function that takes four inputs:

1.  `start_value`
2.  `end_value`
3.  `years` (optional)
4.  `method` (defaults to "simple")

### **Method selection:**

-   `if (method == "simple")`

    If user chose "simple" method

-   `growth_rate <- (end_value - start_value) / start_value * 100`

    Calculates basic percentage change

-   `else if (method == "cagr")`

    If user chose "cagr" method

-   `if (is.null(years)) stop("Need years for CAGR")`

```         
Checks if years was provided; stops with error message if not
```

-   `growth_rate <- ((end_value / start_value)^(1/years) - 1) * 100`

```         
Calculates compound annual growth rate
```

-   `else`

    If user provided any method other than "simple" or "cagr"

-   `stop("Method must be 'simple' or 'cagr'")`

```         
Stops function and shows error with valid options
```

### **Return organized results:**

-   `return(growth_rate)`

    Sends back just the calculated growth rate number (not a list)
:::

::: callout-tip
## Function arguments

1.  **Mandatory arguments**

    -   Must be provided when calling the function
    -   Defined in the function signature without a default value

    ```{r}
    #| label: mandatory-argument
    #| eval: false

    my_function <- function(x, y) {
          return(x + y)
        }
    my_function(3, 5) # x and y are mandatory
    ```

2.  **Optional arguments**

    -   Default value assigned in the function definition
    -   If user does not provide a value, the default value is used.
    -   If user provides a value, it overrides the default
    -   If the default value is `NULL`, the argument is not always necessary

    ```{r}
    #| label: optional-argument
    #| eval: false

    my_function <- function(x, y = 10) {
      return(x + y)
    }
    my_function(3)     # Uses default y = 10
    my_function(3, 5)  # Overrides default, y = 5
    ```

3.  **Argument matching**

    -   **Positional Matching:** Arguments matched to parameters in the function definition based on their order

        ➔ Default behavior when argument names not explicitly used

    -   **Named Matching:** Arguments are explicitly named when calling the function (e.g., `argument_name = value`)

        ➔ Allows for passing arguments in any order and improves code readability

4.  **The `...` argument**

    -   Allows a function to accept additional arguments that are passed on to other functions within its body

        ➔ Useful for creating wrapper functions or functions that need to pass options to underlying methods

    ```{r}
    #| label: ellipse-argument
    #| eval: false

    my_wrapper <- function(data, ...) {
          plot(data, ...) # Pass additional arguments to plot()
        }
    my_wrapper(iris$Sepal.Length, col = "blue", pch = 16)
    ```

5.  **Checking for Missing Arguments**

    -   The missing() function can be used within a function to check if a specific argument was provided by the user or if its default value is being used.

    ```{r}
    #| label: missing-argument
    #| eval: false

    my_function <- function(x, y = 10) {
      if (missing(y)) {
        message("y was not provided, using default.")
      }
      return(x + y)
    }
    ```
:::

## Apply to Multiple Variables

**Class question:** How can we use our function to calculate both the simple and CAGR rates for both GDP and employment with a `for` loop?

```{r}
#| label: we-do-apply

# Here's our data
economic_data

# Here's our function
calculate_growth

# Define which columns to loop through
# columns_to_analyze <- [FILL THIS IN]

# Calculate the number of periods (years) for CAGR
# It's the total number of years minus one
# num_periods <- [FILL THIS IN]

# Loop through each column name
# for ([FILL THIS IN]) {
#   
#   # Get the first value of the current column
#   start_value <- economic_data[FILL THIS IN]
#   
#   # Get the last value of the current column
#   end_value <- economic_data[FILL THIS IN]
#   
#   # Calculate simple growth
#   simple_rate <- [FILL THIS IN]
#   
#   # Calculate compound annual growth rate (CAGR)
#   cagr_rate <- [FILL THIS IN]
#   
#   # Print the results for the current column
#   cat(paste0("--- Results for '", [FILL THIS IN], "' ---\n"))
#   cat(paste0("Simple Growth: ", round(simple_rate, 2), "%\n"))
#   cat(paste0("CAGR: ", round(cagr_rate, 2), "%\n\n"))
# }
```

::: callout-tip
## Extracting Elements

The double-bracket `[[ ]]` is used to extract the content of a single element from a list or data frame.

Think of it as reaching into a box and pulling out the item inside. 📦➡️🍎

**Key Distinction: `[[ ]]` vs. `[ ]`**

The single-bracket `[ ]` **subsets** an object, and the result is always the same type as the original. The double-bracket `[[ ]]` **extracts** an element, and the result is the object *inside* that element.

An easy analogy is a **train vs. a passenger**:

-   `data[1]` is like selecting the first train car. You still have a (smaller) train. 🚂

-   `data[[1]]` is like taking the passenger *out* of the first train car. You now have a person, not a train. 🧍
:::

::: {.callout-note collapse="true"}
## Apply Function Solution

```{r}
#| label: we-do-apply-solution

# Define which columns to loop through
columns_to_analyze <- c("gdp", "employment")

# Calculate the number of periods (years) for CAGR
# It's the total number of years minus one
num_periods <- length(economic_data$year) - 1

# Loop through each column name
for (col in columns_to_analyze) {
  
  # Get the first value of the current column
  start_value <- economic_data[[col]][1]
  
  # Get the last value of the current column
  end_value <- economic_data[[col]][length(economic_data[[col]])]
  
  # Calculate simple growth
  simple_rate <- calculate_growth(start_value, end_value, method = "simple")
  
  # Calculate compound annual growth rate (CAGR)
  cagr_rate <- calculate_growth(start_value, end_value, years = num_periods, method = "cagr")
  
  # Print the results for the current column
  cat(paste0("--- Results for '", col, "' ---\n"))
  cat(paste0("Simple Growth: ", round(simple_rate, 2), "%\n"))
  cat(paste0("CAGR: ", round(cagr_rate, 2), "%\n\n"))
}
```
:::

# Hands-On Practice (`You Do`)

## Core Task: Economic Calculator Functions

Create two functions for common economic calculations using the following practice data:

```{r}
#| label: you-do-setup

# Practice data
set.seed(789)
indicators <- data.frame(
  country = rep(c("USA", "Germany", "Japan"), each = 6),
  year = rep(2015:2020, 3),
  gdp_growth = c(rnorm(6, 2.2, 0.8), rnorm(6, 1.8, 0.6), rnorm(6, 1.1, 0.4)),
  inflation = c(rnorm(6, 2.1, 0.7), rnorm(6, 1.5, 0.5), rnorm(6, 0.8, 0.3))
)

head(indicators, 9)
```

### Core 1: Calculate Volatility Function

**Write a function called `calculate_volatility()` that:**

-   Takes: data, country name, variable name
-   Returns: standard deviation of that variable for that country

```{r}
#| label: you-do-volatility
#| eval: false

calculate_volatility <- function(data, country_name, variable_name) {
  # Your code here
  
}

# Test it
vol_result <- calculate_volatility(indicators, "USA", "gdp_growth")
```

::: callout-tip
## Referring to variables within a function with `dplyr`

The curly-curly `{{ }}` operator (also called the "embrace" operator) is a special tool for making your functions more user-friendly and intuitive. It lets your function accept **bare, unquoted column names**, just like `filter()` or `select()` do.

Think of `{{ }}` as creating a **placeholder** in your function. It tells `dplyr`, "Wait here! The user will give you a bare column name to put in this spot."

```{r}
#| label: bare-names
#| eval: false

# The function expects 'variable_name' to be a bare name, like gdp_per_capita
calculate_mean_embrace <- function(df, country_name, variable_name) {
  df %>%
    filter(country == country_name) %>%
    # We "embrace" the variable that will hold the bare column name
    summarise(mean_value = mean({{ variable_name }}, na.rm = TRUE))
}

# No quotes! It feels just like using summarise() interactively.
calculate_mean_embrace(my_data, "USA", gdp_per_capita)
```
:::

### Core 2: Country Summary Function

**Write a function called `summarize_country()` that:**

-   Takes: data, country name
-   Calculates: mean and SD for both gdp_growth and inflation
-   Prints: a nice summary

```{r}
#| label: you-do-summary
#| eval: false

summarize_country <- function(data, country_name) {
  # Your code here
  
}

# Test it
summarize_country(indicators, "Germany")
```

### Core 3: Apply Your Functions

**Use a `for` loop to:**

1.  Calculate GDP growth volatility for all countries
2.  Create summaries for all countries

```{r}
#| label: you-do-apply
#| eval: false

# Get a list of all unique countries
all_countries <- unique(indicators$country)

# Your for loop here
```

**Success criteria:**

-   Both functions work without errors
-   For loop applies them to all countries
-   Output is clear and informative

## Extension Activities

### Extension 1: Try `lapply`

If you finish early, rewrite your `for` loop using `lapply()`:

```{r}
#| label: you-do-extension
#| eval: false

# Use lapply instead of for loop
volatility_results <- lapply(countries, function(country) {
  calculate_volatility(indicators, country, "gdp_growth")
})
names(volatility_results) <- countries
```

## Working Solutions

Here are the solutions—but seriously, don’t peek until you’ve wrestled with the problems. The struggle is where the learning happens.

::: {.callout-note collapse="true"}
## Core Task Solutions

### Core 1: Volatility Function

```{r}
#| label: core1-solution

calculate_volatility <- function(data, country_name, variable_name) {
  volatility <- data %>%
    filter(country == country_name) %>%
    summarize(volatility = sd({{variable_name}}, na.rm = TRUE)) %>%
    pull(volatility)
  
  return(volatility)
}

# Test
calculate_volatility(indicators, "USA", gdp_growth)
```

**Note:** Since I used the `{{}}` operator in my function, it expects an unquoted variable name, like `gdp_growth` rather than `"gdp_growth"`. If I would like to use a concatenate and print function (`cat`) for detailing the results within my function, I need to turn my bare variable name into a string for printing.

```{r}
#| label: core1-solution2

calculate_volatility <- function(data, country_name, variable_name) {
  volatility <- data %>%
    filter(country == country_name) %>%
    summarize(volatility = sd({{variable_name}}, na.rm = TRUE)) %>%
    pull(volatility)
  
  # Capture the bare variable name and turn it into a string for printing
  var_name_string <- deparse(substitute(variable_name))
  cat("=== Volatility for", country_name, "===\n")
  cat(country_name, var_name_string, "Volatility:", round(volatility, 3), "\n")
  cat("\n") # Add a blank line for spacing
  
  return(volatility)
}

# Test
usa_gdp <- calculate_volatility(indicators, "USA", gdp_growth)
```

### Core 2: Summary Function

```{r}
#| label: core2-solution

summarize_country <- function(data, country_name) {
  # Filter the data and then summarize in a single step
  summary_results <- data %>%
    filter(country == country_name) %>%
    summarize(
      gdp_mean = mean(gdp_growth, na.rm = TRUE),
      gdp_sd = sd(gdp_growth, na.rm = TRUE),
      inf_mean = mean(inflation, na.rm = TRUE),
      inf_sd = sd(inflation, na.rm = TRUE)
    )

  # Use cat() to print the results with rounded numbers
  cat("=== Summary for", country_name, "===\n")
  cat("GDP Growth: Mean =", round(summary_results$gdp_mean, 2), 
      "SD =", round(summary_results$gdp_sd, 2), "\n")
  cat("Inflation: Mean =", round(summary_results$inf_mean, 2), 
      "SD =", round(summary_results$inf_sd, 2), "\n\n")
  
}

# Test
summarize_country(indicators, "Germany")
```

### **Core 3: `for` Loop**

```{r}
#| label: core3-solution

# Get a list of all unique countries
all_countries <- unique(indicators$country)

# Loop to Calculate Volatility & Create Summaries
for (country in all_countries) {
  calculate_volatility(indicators, country, gdp_growth)
  summarize_country(indicators, country)
}

```
:::

::: {.callout-note collapse="true"}
## Extension Activity Solutions

### Extension 1: `lapply`

```{r}
#| label: extension1-solution

countries <- unique(indicators$country)

# Use lapply with an anonymous (lambda) function to run both commands
# We wrap it in invisible() because we only care about the printing from cat(),
#   not the list of volatility values that lapply returns.
invisible(lapply(all_countries, function(nation) {
  # This temporary function runs for each country
  calculate_volatility(indicators, nation, gdp_growth)
  summarize_country(indicators, nation)
}))
```
:::

# Wrap-Up & Looking Ahead

## What We Learned Today

**Functions:**

-   DRY principle: Don't repeat yourself
-   Write once, use everywhere
-   Test with simple inputs, add debugging prints

**Basic Iteration:**

-   `for` loops: explicit and reliable
-   `lapply`/`sapply`: functional programming
-   `apply`: for matrices (econometric contexts)

## Thursday Preview: APIs + Advanced Iteration

Thursday we'll learn APIs (getting data from the web) and discover why `purrr::map()` is perfect for API workflows. APIs naturally create repetitive tasks—exactly when elegant iteration shines!

**Teaser:** Instead of manually downloading unemployment data for 50 states:

```{r}
#| label: api-teaser
#| eval: false

# Coming Thursday!
get_state_unemployment <- function(state_code) {
  # Query Bureau of Labor Statistics API
  # Return clean data
}

# Get all states automatically  
states <- c("CA", "TX", "NY", ...)
unemployment_data <- map_dfr(states, get_state_unemployment)
```

## Reading for Thursday

**Required reading:**

-   [**An intro to APIs**](https://zapier.com/resources/guides/apis) **(Chapters 1–4)**

**Suggested:**

-   **R4DS Chapter 23 (Hierarchical Data)**
-   `purrr::map()` [tutorial](https://jennybc.github.io/purrr-tutorial/ls01_map-name-position-shortcuts.html)
-   `purrr` [cheatsheet](https://rstudio.github.io/cheatsheets/html/purrr.html?_gl=1*k1z7fk*_ga*MTkzMzExNjg1OC4xNzU0NDkwMDc5*_ga_2C0WZ1JHG0*czE3NTQ0OTAwNzkkbzEkZzEkdDE3NTQ0OTAwODkkajUwJGwwJGgw)
-   `httr2` [get started](https://httr2.r-lib.org/articles/httr2.html)
-   `httr2` [wrapping APIs](https://httr2.r-lib.org/articles/wrapping-apis.html)
