---
title: "Lecture 6: Data Acquisition - APIs & Iteration"
subtitle: "APEC 8221: Programming for Econometrics"
author: "Ali Joglekar"
date: "September 18, 2025"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: show
    code-tools: true
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
  eval: true
  echo: true
editor: visual
---

# Opening & Check-In

## Recap: Functions from Tuesday

**Quick reminder:**

-   Writing custom functions to eliminate repetitive code
-   DRY principle: "Don't Repeat Yourself"
-   Functions have arguments, body, and return values
-   Testing functions with simple inputs first

Any lingering questions about function writing?

## Today's Mission

Today we're tackling **programmatic data acquisition** using APIs (Application Programming Interfaces). But here's the thing—APIs are perfect for demonstrating why iteration matters.

**Real scenario:** You need quarterly GDP data for 20 countries from the Federal Reserve's database. You could manually click through 20 web pages and copy-paste data... or you could write 5 lines of R code that does it automatically.

**Why this combination matters:**

-   APIs naturally create repetitive tasks (same request, different parameters)
-   You'll immediately see why `purrr::map()` is essential
-   Both APIs and iteration are core skills for modern data acquisition

By the end of today, you'll be pulling economic data from the web like a pro, automatically.

# What Are APIs?

## The Big Picture

An **API (Application Programming Interface)** is like a waiter at a restaurant:

-   **You** (the customer) want data
-   **The API** (the waiter) takes your request to the kitchen
-   **The database** (the kitchen) prepares your data
-   **The API** brings back exactly what you asked for

An API is just a contract:

-   You follow the provider’s directions
-   The provider commits to giving you data in a pre-specified format.

APIs are everywhere in economic research:

-   **FRED** (Federal Reserve Economic Data) – US economic indicators
-   **World Bank** – International development data
-   **Bureau of Labor Statistics** – Employment, inflation, wages
-   **Census Bureau** – Demographics, business patterns

### APIs vs. Web Scraping: Choosing the Right Tool

| **When to Use APIs** | **When to Use Web Scraping** |
|----|----|
| ✅ Data provider offers an API | ✅ No API available for the data you need |
| ✅ Need real-time or frequently updated data | ✅ Data is publicly displayed but not via API |
| ✅ Large-scale, repeated data collection | ✅ One-time or small-scale data collection |
| ✅ Provider encourages programmatic access | ✅ Website allows scraping (check robots.txt) |
| ✅ Structured, consistent data format | ✅ Data embedded in web pages |

**Bottom line:** Use the tool that fits your data source and research needs. Both have legitimate uses in modern data acquisition.

### General API Structure

Every API consists of a `URL` to a **domain** and a path to an **endpoint**.

-   **API endpoint:** point at which API connects with the software program
-   APIs work by sending requests for information from a web application or web server and receiving a response

### **Formula for API Calls**

Most professional API work follows this pattern—you build exactly the URL you need for exactly the data you want:

1.  Construct Base URL
2.  Construct Request Parameter String
3.  Construct Full URL-Based Query
4.  Send Query to the Endpoint
5.  Convert Resulting Data from JSON to R-Based Object

## Class Example: FRED API

The Federal Reserve Economic Data (FRED) API gives us access to 800,000+ economic time series. Perfect for economics research!

**What we can get:**

-   GDP, unemployment, inflation rates
-   Interest rates, money supply
-   International trade data
-   Regional economic indicators

Let's start with a single series, Gross Domestic Product, and then see why we need iteration.

# Single API Request Demo (`I Do`)

### Step 0: Obtain API Key (`We Do`)

First we need to get an API key.

-   Navigate to the St. Louis FED FRED API key webpage: <https://fred.stlouisfed.org/docs/api/api_key.html>

-   Sign in or create an account

-   Click the [Request or view your API keys](https://fredaccount.stlouisfed.org/apikeys) link

-   Click the `+ Request API Key` button and fill out the requested information

-   Copy the API key number and save it in your R environment for future reference

    ```{r}
    #| label: save-key


    # Save FRED API key in .Renviron file, then Restart R
    usethis::edit_r_environ()

    # Store it in .Renviron as FRED_API_KEY = "your_key_here"
    ```

-   Restart R \[Session \> Restart R\] – the shortcut is `Ctrl + Shift + F10` on a Windows PC and `Command + Shift + F10` on a Mac

Now you can load your API key into your workspace for the lecture

```{r}
#| label: load-api-key

# Secure API key storage – load from .Renviron file
fred_key <- Sys.getenv("FRED_API_KEY")


# For demo purposes, checking if key is available
if (fred_key == "") {
  cat("Note: FRED API key not found in .Renviron file\n")
  cat("Get a free key at: https://fred.stlouisfed.org/docs/api/api_key.html\n")
  fred_key <- "demo_key_placeholder"
}
```

### Step 1: Construct Base URL

```{r}
#| label: base-url

# Step 1: Construct Base URL for FRED API
url_domain <- "https://api.stlouisfed.org/fred/"    # The API's address
url_endpoint <- "series/observations"               # What type of data we want
url_base <- paste0(url_domain, url_endpoint)        # Combine the two

cat("Base API URL:", url_base, "\n")
```

### Step 2: Construct Request Parameter String

This approach is standard across virtually all APIs—you specify exactly what you want rather than downloading everything and filtering later.

```{r}
#| label: parameter-string

# Parameters (what specific data you want -- filters)
# These get added after a "?" and separated by "&"
parameters <- list(
  series_id = "GDP",                  # Which economic indicator?
  api_key = fred_key,                 # Your authentication
  file_type = "json",                 # What format?
  observation_start = "2020-01-01",   # What time period?
  observation_end = "2023-12-31"
)

# Parameter query string: 
param_string <- paste0(names(parameters), "=", parameters, collapse = "&")

cat("Parameter Query String:", param_string, "\n")
```

### Step 3: **Construct Full URL-Based Query Manually**

```{r}
#| label: full-url-query

url_query <- paste(url_base, param_string, sep = '?')

cat("Full URL-Based Query:", url_query, "\n")
```

### Steps 2-3: **Construct Full URL-Based Query with `httr2`**

We can automate steps 2-3 using the `httr2` package.

```{r}
#| label: build-httr-request

# Our toolkit for today
library(tidyverse)
library(httr2)      # for making web requests
library(jsonlite)   # for easy parsing of JSON responses

# Step 2a: Set up a request to the base URL
gdp_url <- request(url_base) %>%
  
  # Step 2b: Add our specific parameters (what data we want) using req_url_query()
  req_url_query(
    series_id = "GDP",        # We want GDP data specifically
    api_key = fred_key,       # Our authentication key
    file_type = "json"        # We want JSON format (easier to parse)
  )

gdp_url
```

### Step 4: Send Query to the Endpoint

```{r}
#| label: make-request

# Make the request using req_perorm()
gdp_response <- gdp_url %>% req_perform()

# Check if it worked (status code 200 = success)
cat("Status code:", resp_status(gdp_response), "\n")
if (resp_status(gdp_response) == 200) {
  cat("✅ Success! Data retrieved.\n")
} else {
  cat("❌ Something went wrong.\n")
}
```

### Step 5: Parse the JSON Response

::: callout-tip
## Common API Response Format: JSON

JavaScript Object Notation (JSON) is a javascript-based way of specifying structured data.

-   It is centered around name value pairs
-   Values can contain vectors in \[\]
-   Objects are contained in {}
-   Items in a vector or object are separated by ,
-   Text is enclosed in “ ”

``` json
{"class":"APEC8991",
 "students":[{"name":"John Doe", "grade":70},
             {"name":"Jane Smith", "grade":99}]
}
```

The `jsonlite` package has a handy `fromJSON()` function, where you give it a string of JSON text, and it gives you back an R list

``` json
{"class":"APEC8991",
 "students":[{"name":"John Doe", "grade":70},
             {"name":"Jane Smith", "grade":99}]
}
```

becomes

``` r
list("class"="APEC8991", 
     "students"=c(list("name"="John Doe", "grade"=70), 
                  list("name"="Jane Smith", "grade"=99)))
```
:::

```{r}
#| label: parse-json

# Convert resulting data from JSON to R-based object using resp_body_json()
gdp_raw <- gdp_response %>% resp_body_json()

# Look at the structure - this helps us understand how to extract data
str(gdp_raw, max.level = 2)

# The actual data is usually nested inside the JSON structure
# FRED puts the time series data in the "observations" section
gdp_observations <- gdp_raw$observations

# Look at first few observations to understand the data format
head(gdp_observations, 3) %>% str()
```

### Step 6: Convert to Tidy Data

```{r}
#| label: tidy-data

# Extract observations element from list -- this is where our data is
gdp_observations <- gdp_raw$observations

# Look at the structure of our data
glimpse(gdp_observations)

# We have a list of data frames -- not really helpful
# Use dplyr::bind_rows to convert a list of data frames into a single data-frame
# bind_rows() combines each list element vertically (row-wise)
gdp_data <- bind_rows(gdp_observations)
gdp_data

# Clean up our data
gdp_data <- gdp_data %>% 
  # Select columns needed
  select(date, value) %>% 
  # Format date & value appropriately
  # Add series id
  mutate(date = as.Date(date), 
         value = as.numeric(value), 
         series_id = "GDP") %>% 
  # Remove missing values
  filter(!is.na(value))


dates <- character(length(gdp_observations))
values <- numeric(length(gdp_observations))

# Extract data from each observation
for (i in seq_along(gdp_observations)) {
  dates[i] <- gdp_observations[[i]]$date
  values[i] <- as.numeric(gdp_observations[[i]]$value)
}

# Look at our clean data
head(gdp_data)
tail(gdp_data)

# Quick plot to verify
ggplot(gdp_data, aes(x = date, y = value)) +
  geom_line(color = "steelblue", linewidth = 1) +
  labs(
    title = "US Gross Domestic Product",
    x = "Year",
    y = "GDP (Billions of dollars)",
    caption = "Source: Federal Reserve Economic Data (FRED)"
  ) +
  theme_minimal()
```

Great! We have one economic time series. But what if we need multiple series?

## The Repetition Problem

Suppose we want these 5 economic indicators:

-   **GDP** - Gross Domestic Product
-   **UNRATE** - Unemployment Rate
-   **CPIAUCSL** - Consumer Price Index
-   **FEDFUNDS** - Federal Funds Rate -
-   **DEXUSEU** - US/Euro Exchange Rate

**The manual approach:**

```{r}
#| label: manual-approach
#| eval: false

# This is tedious and error-prone...
gdp_url <- request(url_base) %>%
  req_url_query(series_id = "GDP", api_key = fred_key, file_type = "json")
gdp_response <- gdp_url %>% req_perform()
gdp_data <- process_response(gdp_response)  # Imagine this function exists

unrate_url <- request(url_base) %>%
  req_url_query(series_id = "UNRATE", api_key = fred_key, file_type = "json")
unrate_response <- unrate_url %>% req_perform()
unrate_data <- process_response(unrate_response)

# ... repeat 3 more times ...
# This is exactly what functions and iteration solve!
```

There has to be a better way... and there is!

## Enter Iteration with `purrr`

### The Function Approach

First, let's wrap our API workflow in a function:

```{r}
#| label: api-function

# Now let's wrap our API workflow in a reusable function
# This function takes a series ID and returns clean data

get_fred_series <- function(series_id, api_key = fred_key) {
  # Always let the user know what's happening
  cat("Fetching", series_id, "...\n")
  
  # Step 1: Build and execute the request (same as before)
  response <- request(url_base) %>%
    req_url_query(
      series_id = series_id,    # This changes for each function call
      api_key = api_key,        # This stays the same
      file_type = "json"        # This stays the same
    ) %>%
    req_perform()
  
  # Step 2: Check if the request worked (professional habit!)
  if (resp_status(response) != 200) {
    warning("Failed to get data for ", series_id)
    return(NULL)  # Return nothing if it failed
  }
  
  # Step 3: Parse the JSON (extract the data we want)
  raw_data <- response %>% resp_body_json()
  
  # Step 4: Convert to clean, tidy data frame
  # Extract observations and convert to simple data frame
  observations <- raw_data$observations
  
  # Create vectors for date and value
  dates <- character(length(observations))
  values <- numeric(length(observations))
  
  for (i in seq_along(observations)) {
    dates[i] <- observations[[i]]$date
    values[i] <- as.numeric(observations[[i]]$value)
  }
  
  # Step 5: Build final data frame
  clean_data <- tibble(
    date = as.Date(dates),
    value = values,
    series_id = series_id
  ) %>%
    filter(!is.na(value))  # Remove missing observations
  
  # Let us know what happened
  cat("✅ Got", nrow(clean_data), "observations for", series_id, "\n")
  return(clean_data)
}

# Test our function - this should give us clean GDP data
test_data <- get_fred_series("GDP")
head(test_data)
```

### The Power of `map()`: Understanding What's Happening

Now we can get multiple series efficiently. But let's be clear about what `map()` is actually doing:

```{r}
#| label: map-demo

# Our list of economic indicators
economic_series <- c("GDP", "UNRATE", "CPIAUCSL", "FEDFUNDS", "DEXUSEU")

# What map() does step by step:
# Step 1: Call get_fred_series("GDP") -> returns cleaned GDP data
# Step 2: Call get_fred_series("UNRATE") -> returns cleaned unemployment data  
# Step 3: Call get_fred_series("CPIAUCSL") -> returns cleaned CPI data
# Step 4: Call get_fred_series("FEDFUNDS") -> returns cleaned fed funds data
# Step 5: Call get_fred_series("DEXUSEU") -> returns cleaned exchange rate data
# Step 6: Combine all results into one big data frame

# The manual way (to understand what we're automating):
all_results <- list()
for (i in seq_along(economic_series)) {
  all_results[[i]] <- get_fred_series(economic_series[i])
}
# Combine results
all_data_manual <- bind_rows(all_results)

# The map() way - exact same result, cleaner code:
all_data_map <- economic_series %>%
  map(get_fred_series) %>%  # Apply function to each series
  bind_rows()               # Combine all results

# Look at what we got
glimpse(all_data_map)

# Quick summary
all_data_map %>%
  group_by(series_id) %>%
  summarise(
    observations = n(),
    start_date = min(date),
    end_date = max(date),
    .groups = "drop"
  )
```

**Key Point:** `map()` is making 5 separate API calls behind the scenes, one for each series. It's NOT building one giant URL with all 5 series—that's not how most APIs work.

**This is exactly how professional data scientists work with APIs:**

1.  Write a function that gets exactly what you need from one API call
2.  Use `map()` to repeat that function across multiple parameters
3.  Let `map()` handle the repetition and combining

**We just retrieved 5 economic time series with clean, efficient code!**

## Professional API Workflow

Let's add some real-world robustness:

```{r}
#| label: professional-workflow

get_fred_series_robust <- function(series_id, api_key = fred_key, delay = 1) {
  cat("Fetching", series_id, "...\n")
  
  # Respect rate limits
  Sys.sleep(delay)
  
  # Error handling
  tryCatch({
    response <- request(url_base) %>%
      req_url_query(
        series_id = series_id,
        api_key = api_key,
        file_type = "json"
      ) %>%
      req_perform()
    
    if (resp_status(response) != 200) {
      warning("API returned status ", resp_status(response), " for ", series_id)
      return(NULL)
    }
    
    raw_data <- response %>% resp_body_json()
    observations <- raw_data$observations
    
    # Simple data extraction
    dates <- character(length(observations))
    values <- numeric(length(observations))
    
    for (i in seq_along(observations)) {
      dates[i] <- observations[[i]]$date
      values[i] <- as.numeric(observations[[i]]$value)
    }
    
    clean_data <- tibble(
      date = as.Date(dates),
      value = values,
      series_id = series_id
    ) %>%
      filter(!is.na(value))
    
    cat("✅ Success:", nrow(clean_data), "observations\n")
    return(clean_data)
    
  }, error = function(e) {
    warning("Error fetching ", series_id, ": ", e$message)
    return(NULL)
  })
}

# Test with delay and error handling
robust_data <- economic_series %>%
  map(get_fred_series_robust, delay = 0.5) %>%
  bind_rows()

nrow(robust_data)
```

# Debugging Interlude: API Troubleshooting

Let's see the most common API problems and how to fix them:

```{r}
#| label: debugging-apis
#| eval: false

# Problem 1: Rate limiting (HTTP 429)
# "Too Many Requests" or status code 429
# Solution: Add Sys.sleep(1) between requests

for(i in 1:5) {
  response <- request(url_base) %>% req_perform()
  Sys.sleep(1)  # Wait 1 second between requests
}

# Problem 2: Authentication errors (HTTP 401) 
# "Unauthorized" or status code 401
# - Check your API key spelling
# - Make sure API key hasn't expired
# - Check if you need to register key

# Problem 3: Bad requests (HTTP 400)
# "Bad Request" or status code 400
# - Check parameter names and values
# - Look at API documentation again
# - Try a simpler request first

# Always check status codes before parsing:
response <- request(url_base) %>% req_perform()

if (resp_status(response) == 200) {
  # Safe to parse data
  data <- response %>% resp_body_json()
} else {
  # Handle the error
  stop("API request failed with status: ", resp_status(response))
}

# Debugging JSON parsing:
# Look at raw response first, then parse structure
raw_content <- response %>% resp_body_string()
print(raw_content)  # See the raw JSON

parsed <- response %>% resp_body_json()
str(parsed)  # See the structure
```

**Your API debugging workflow:** 1. **CHECK** status codes before parsing 2. **EXAMINE** raw response when confused 3. **START** with simple requests 4. **RESPECT** rate limits (add delays)

# Guided Practice: Building Together (`We Do`)

Let's work through converting a manual API workflow to an automated one.

## The Challenge

I want to compare economic performance across different regions. I need these FRED series:

-   **GDP** - US Total
-   **CANGDP** - California GDP
-   **TXGDP** - Texas GDP
-   **NYGDP** - New York GDP

Let me start manually, and you help me convert it to use `map()`:

```{r}
#| label: we-do-manual
#| eval: false

# Manual approach (tedious!)
us_gdp <- get_fred_series("GDP")
ca_gdp <- get_fred_series("CAGDP") 
tx_gdp <- get_fred_series("TXGDP")
ny_gdp <- get_fred_series("NYGDP")

# Manually combine
regional_data <- bind_rows(us_gdp, ca_gdp, tx_gdp, ny_gdp)
```

**Class discussion:**

1.  What's repetitive about this code?
2.  How can we use a vector and `map()` to simplify this?
3.  What should our vector contain?

**Let's build the solution together:**

```{r}
#| label: we-do-solution

# What we decided as a class:
regional_series <- c("GDP", "CAGDP", "TXGDP", "NYGDP")

# Convert to map() approach
regional_data <- regional_series %>%
  map(get_fred_series_robust, delay = 0.5) %>%
  bind_rows()

# Quick verification
regional_data %>%
  group_by(series_id) %>%
  summarise(
    observations = n(),
    latest_date = max(date),
    latest_value = last(value),
    .groups = "drop"
  )
```

**Discussion:** What are the advantages of the `map()` approach over manual repetition?

# Hands-On Practice: Your Turn (`You Do`)

## Core Task: API Parameter Practice (25 minutes)

Now you'll practice the key API skills: building URLs with different parameters and using functions for automation.

### Task 1: Modify API Parameters (Single Request)

```{r}
#| label: core-task-1
#| eval: false

# Practice changing API parameters to get different data
# Get unemployment data for just the years 2020-2023

unemployment_url <- request(url_base) %>%
  req_url_query(
    series_id = "UNRATE",           # Unemployment rate
    api_key = fred_key,
    file_type = "json",
    observation_start = "___",      # Fill in start date
    observation_end = "___"         # Fill in end date
  )

unemployment_response <- unemployment_url %>% req_perform()

# Check if it worked and look at the data
if (resp_status(unemployment_response) == 200) {
  cat("✅ Success!\n")
  raw_data <- unemployment_response %>% resp_body_json()
  cat("Got", length(raw_data$observations), "observations\n")
} else {
  cat("❌ Failed with status:", resp_status(unemployment_response), "\n")
}
```

### Task 2: Enhanced Function with Multiple Parameters

```{r}
#| label: core-task-2
#| eval: false

# Create a function that lets you control the date range
get_fred_series_dates <- function(series_id, start_date, end_date, api_key = fred_key) {
  cat("Fetching", series_id, "from", start_date, "to", end_date, "\n")
  
  # Build your request using the parameters
  response <- request(url_base) %>%
    req_url_query(
      series_id = series_id,
      api_key = api_key,
      file_type = "json",
      observation_start = start_date,
      observation_end = end_date
    ) %>%
    req_perform()
  
  # Add error checking and data processing here
  # (Use the same pattern from our earlier function)
  
  # Your code here...
}

# Test your function
test_data <- get_fred_series_dates("GDP", "2020-01-01", "2023-12-31")
```

### Task 3: Automate Multiple API Calls

```{r}
#| label: core-task-3
#| eval: false

# Use your new function to get recent data for multiple indicators
indicators <- c("UNRATE", "CPIAUCSL", "HOUST")

# Method 1: Manual approach (understand what we're automating)
unrate_data <- get_fred_series_dates("UNRATE", "2022-01-01", "2024-01-01")
cpi_data <- get_fred_series_dates("CPIAUCSL", "2022-01-01", "2024-01-01")
housing_data <- get_fred_series_dates("HOUST", "2022-01-01", "2024-01-01")

# Method 2: Automated approach with map
all_data <- indicators %>%
  map(~ get_fred_series_dates(.x, "2022-01-01", "2024-01-01")) %>%
  bind_rows()  # Combine results

# Compare what you get - should be the same!
```

## Extension Activities

### Extension 1: API Metadata Exploration

Get ready for pagination by learning to read API metadata:

```{r}
#| label: extension-pagination-prep
#| eval: false

# Make a request and examine the full response structure
response <- request(url_base) %>%
  req_url_query(
    series_id = "UNRATE",
    api_key = fred_key,
    file_type = "json",
    limit = 100  # Only get 100 observations
  ) %>%
  req_perform()

full_data <- response %>% resp_body_json()

# Explore the metadata - what information is available?
str(full_data, max.level = 1)

# How many total observations are available?
# How many did we actually get?
# What parameters could you use for pagination?
```

### Extension 2: Error Handling & Logging

Test your API skills with series that might not exist:

```{r}
#| label: extension-error-handling
#| eval: false

# Some of these series don't exist - can you handle the errors?
test_series <- c("GDP", "UNRATE", "FAKE_SERIES", "INVALID123")

# Modify get_fred_series_dates to return status information
# Challenge: Make it return both success/failure info AND data when available
```

### Extension 3: Parameter Exploration

Try different API parameters to see what's possible:

```{r}
#| label: extension-parameters
#| eval: false

# Experiment with other FRED API parameters:
# - frequency: "a" (annual), "q" (quarterly), "m" (monthly)
# - units: "lin" (levels), "chg" (change), "pch" (percent change)

# Example: Get GDP in percent change format
response <- request(url_base) %>%
  req_url_query(
    series_id = "GDP",
    api_key = fred_key,
    file_type = "json",
    units = "pch"  # Percent change
  ) %>%
  req_perform()

# Compare this to the raw levels data - what's different?
```

## Working Solutions

Here are the solutions—but seriously, don't peek until you've wrestled with the problems!

::: {.callout-note collapse="true"}
## Core Task Solutions

### Task 1 Solution:

```{r}
#| label: solution-task1

unemployment_url <- request(url_base) %>%
  req_url_query(
    series_id = "UNRATE",
    api_key = fred_key,
    file_type = "json",
    observation_start = "2020-01-01",
    observation_end = "2023-12-31"
  )

unemployment_response <- unemployment_url %>% req_perform()

if (resp_status(unemployment_response) == 200) {
  cat("✅ Success!\n")
  raw_data <- unemployment_response %>% resp_body_json()
  cat("Got", length(raw_data$observations), "observations\n")
} else {
  cat("❌ Failed with status:", resp_status(unemployment_response), "\n")
}
```

### Task 2 Solution:

```{r}
#| label: solution-task2

get_fred_series_dates <- function(series_id, start_date, end_date, api_key = fred_key) {
  cat("Fetching", series_id, "from", start_date, "to", end_date, "\n")
  
  response <- request(url_base) %>%
    req_url_query(
      series_id = series_id,
      api_key = api_key,
      file_type = "json",
      observation_start = start_date,
      observation_end = end_date
    ) %>%
    req_perform()
  
  if (resp_status(response) != 200) {
    warning("Failed to get data for ", series_id)
    return(NULL)
  }
  
  raw_data <- response %>% resp_body_json()
  observations <- raw_data$observations
  
  dates <- character(length(observations))
  values <- numeric(length(observations))
  
  for (i in seq_along(observations)) {
    dates[i] <- observations[[i]]$date
    values[i] <- as.numeric(observations[[i]]$value)
  }
  
  clean_data <- tibble(
    date = as.Date(dates),
    value = values,
    series_id = series_id
  ) %>%
    filter(!is.na(value))
  
  cat("✅ Got", nrow(clean_data), "observations\n")
  return(clean_data)
}

# Test the function
test_data <- get_fred_series_dates("GDP", "2020-01-01", "2023-12-31")
head(test_data)
```

### Task 3 Solution:

```{r}
#| label: solution-task3

indicators <- c("UNRATE", "CPIAUCSL", "HOUST")

# Automated approach with map
all_data <- indicators %>%
  map(~ get_fred_series_dates(.x, "2022-01-01", "2024-01-01")) %>%
  bind_rows()

# Summary
all_data %>%
  group_by(series_id) %>%
  summarise(
    observations = n(),
    start_date = min(date),
    end_date = max(date),
    .groups = "drop"
  )
```
:::

# Wrap-Up & Looking Ahead

## What We Learned Today

**API Fundamentals:**

-   APIs are structured ways to request data programmatically -
-   Always check status codes before parsing responses -
-   Respect rate limits and handle errors gracefully

**The Power of Iteration:** -

-   `map()` eliminates repetitive API calls -
-   Functions + `purrr` = scalable data acquisition -
-   Professional workflows include error handling and logging

## Why This Actually Matters

**Dissertation scenarios where you'll use these skills:**

1.  **International Development Research:**
    -   Pull World Bank data for 150+ countries automatically
    -   No more manual Excel downloads and data alignment
2.  **Regional Economics:**
    -   Get county-level employment data for all US counties
    -   Would take weeks manually, takes minutes with APIs
3.  **Financial Economics:**
    -   Download daily stock prices for S&P 500 companies
    -   Update your analysis automatically when new data arrives

**Professional scenarios:** - **Fed Economist:** Update quarterly forecasting models with latest data - **Consulting:** Provide clients with real-time economic dashboards\
- **Academic Research:** Ensure your replication files actually work

## Next Lecture Preview: Reshaping and Joining Data

**Tuesday we tackle messy, real-world data structures:**

-   **Tidy data principles** - when data isn't shaped the way you need
-   **Reshaping** with `pivot_longer()` and `pivot_wider()`
-   **Joining datasets** with `left_join()`, `inner_join()`, etc.

**Why this follows perfectly:** APIs often return data in inconvenient formats. Tuesday you'll learn to reshape and combine that data for analysis.

### Reading for Tuesday

**Required:**

-   **R4DS Chapter 5 (Tidy Data)** - The foundation of data reshaping
-   **R4DS Chapter 19 (Joins)** - Combining multiple datasets
-   **R4DS Chapter 18 (Missing Values)** - Handling real-world data issues

**Suggested:**

-   **tidyr cheatsheet** - Keep this handy for reshaping
-   **ERP 6.2-6.5 (Data Carpentry)** - Professional data cleaning workflows

## Assignment 2 Posted Friday

**Assignment 2: Automation and API Data Acquisition**

-   **Posted:** Friday, September 19 at 5 PM (or as close to that time as I can get)

-   **Due:** Saturday, September 27 at 8 AM
