---
title: "APEC 8221 - Assignment 1: Tidyverse Fundamentals"
author: "[Runcheng Xu]"
date: Sep 19 2025
format: 
  pdf:
    number-sections: true
    colorlinks: true
    embed-resources: true
execute:
  warning: false
  message: false
  eval: true
  echo: true
editor: visual
---

```{=html}
<!-- 
Before you begin:
1. Save this file as assignment1-[your-x500-id].qmd
2. Replace "[REPLACE WITH YOUR NAME]" above with your actual name
3. Make sure you're working in your course R Project
4. Include your GitHub repository URL in Canvas submission
-->
```

**GitHub Repository URL:** \[<https://github.com/Runcheng112/apec8221-xu001296.git>\]

# Setup

```{r}
#| label: setup

# Load required packages
library(tidyverse)
library(gapminder)

# Check data structure
glimpse(gapminder)
```

# Data Analysis & Visualization Practice

## Top Performers Identification

Find the top 3 countries in each continent for highest life expectancy and highest GDP per capita in 2007.

### Highest Life Expectancy by Continent (2007)

```{r}
#| label: top-life-exp
library(dplyr)

# Top 3 by Life Expectancy 
top_lifeExp <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  arrange(desc(lifeExp), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  select(continent, country, lifeExp)

# 打印结果
print(top_lifeExp, n = Inf)


# Your code here - find top 3 countries by life expectancy in each continent
# Remember to use group_by(), arrange(), and slice_head()

```

In 2007, countries in Europe and Asia showed the highest life expectancies, with places like Hong Kong, Japan, and Iceland exceeding 82 years.

In contrast, African countries such as Réunion, Libya, and Tunisia had much lower life expectancies, around 76 years, highlighting the health and living standard disparities across continents.

### Highest GDP per Capita by Continent (2007)

```{r}
#| label: top-gdp

# Top 3 by GDP per capita (2007)
top_gdpPercap <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  arrange(desc(gdpPercap), .by_group = TRUE) %>%
  slice_head(n = 3) %>%
  select(continent, country, gdpPercap)

print(top_gdpPercap, n = Inf)
   
# Your code here - find top 3 countries by GDP per capita in each continent

```

In 2007, the highest GDP per capita countries were mostly in Europe, Asia, and the Americas, with Norway, Kuwait, and the United States leading.

African countries like Gabon, Botswana, and Equatorial Guinea also appeared in the top three for the continent, though their GDP per capita was still lower compared to wealthier regions.

## Development Progress Visualization

Create a scatter plot comparing each country's 1952 vs. 2007 life expectancy.

```{r}
#| label: life-exp-data-prep
library(dplyr)
library(ggplot2)

lifeExp_1952 <- gapminder %>%
  filter(year == 1952) %>%
  select(country, continent, lifeExp_1952 = lifeExp)

lifeExp_2007 <- gapminder %>%
  filter(year == 2007) %>%
  select(country, lifeExp_2007 = lifeExp, pop_2007 = pop)

lifeExp_data <- left_join(lifeExp_1952, lifeExp_2007, by = "country")

lifeExp_data <- lifeExp_data %>%
  filter(!is.na(lifeExp_1952), !is.na(lifeExp_2007))

```

```{r}
#| label: life-exp-plot
#| fig-width: 10
#| fig-height: 8

# Create the scatter plot
ggplot(lifeExp_data,
       aes(x = lifeExp_1952, y = lifeExp_2007,
           color = continent, size = pop_2007)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  labs(
    title = "Life Expectancy: 1952 vs 2007",
    x = "Life Expectancy (1952)",
    y = "Life Expectancy (2007)",
    color = "Continent",
    size  = "Population (2007)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
  # Your code here - add the diagonal reference line using geom_abline()
  # Add points with color by continent and size by population
  # Add professional labels and theme

```

**Analysis:** The plot shows that most countries saw clear improvements in life expectancy from 1952 to 2007, since nearly all points are above the diagonal line. While Europe and parts of Asia reached the highest levels, many African countries still lagged behind, though they also made progress.

## Development Progress Classification

Using the `lifeExp_data` from Task 2.1.2, classify countries by improvement level.

```{r}
#| label: life-exp-classification
library(dplyr)

lifeExp_data <- lifeExp_data %>%
  mutate(
    improvement = lifeExp_2007 - lifeExp_1952,
    improvement_level = case_when(
      improvement >= 30 ~ "Major Improvement",
      improvement >= 15 ~ "Moderate Improvement",
      improvement >= 0  ~ "Minor Improvement",
      TRUE              ~ "Decline"
    )
  )

improvement_summary <- lifeExp_data %>%
  group_by(continent, improvement_level) %>%
  summarise(n_countries = n(), .groups = "drop")

print(improvement_summary, n = Inf)


```

### Analysis Questions

**1. Which continent has the most countries with "Major Improvement"?**

```{r}
#| label: major-improvement-analysis

# Your code here - analyze major improvement by continent
major_improvement_summary <- lifeExp_data %>%
  filter(improvement_level == "Major Improvement") %>%
  group_by(continent) %>%
  summarise(n_countries = n(), .groups = "drop") %>%
  arrange(desc(n_countries))

print(major_improvement_summary)

```

Asia is the continent with most "Major Improvement Countries".

**2. Are there any countries that experienced decline? Which ones and why might this have happened?**

```{r}
#| label: decline-analysis

# Your code here - find and examine countries with decline
decline_countries <- lifeExp_data %>%
  filter(improvement_level == "Decline") %>%
  select(country, continent, lifeExp_1952, lifeExp_2007, improvement)

print(decline_countries, n = Inf)

```

In Swaziland and Zimbabwe, life expectancy fell mainly because of the HIV/AIDS epidemic. Economic crisis and weak health systems also made the situation worse, preventing improvements seen in other countries.

**3. What does this tell us about global health convergence?**

The results suggest that global health is partly converging. Many countries in Asia and the Americas made large gains in life expectancy, helping them catch up with Europe. However, Africa shows a mixed picture, with most countries improving only moderately and a few even declining, which means convergence is uneven across regions.

## Economic Development Trajectories

Create a line plot showing life expectancy changes over time for selected countries.

### Step 1: Guided Discovery

```{r}
#| label: trajectory-discovery

life_exp_changes <- lifeExp_data %>%
  mutate(change = lifeExp_2007 - lifeExp_1952) %>%
  select(country, continent, lifeExp_1952, lifeExp_2007, change)

top_gainer <- life_exp_changes %>%
  arrange(desc(change)) %>%
  slice(1)

decliners <- life_exp_changes %>%
  filter(change < 0)

print(top_gainer)
print(decliners)

```

### Step 2: Create Trajectory Visualization

```{r}
#| label: trajectory-plot
#| fig-width: 10
#| fig-height: 6

# Step 2: Create Trajectory Visualization
selected_countries <- c("Oman", "Swaziland", "Zimbabwe", 
                        "Japan", "United States", "United Kingdom")

trajectory_data <- gapminder %>%
  filter(country %in% selected_countries)

ggplot(trajectory_data, aes(x = year, y = lifeExp, color = country)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Life Expectancy Trajectories (1952–2007)",
       x = "Year",
       y = "Life Expectancy",
       color = "Country") +
  theme_minimal()

```

**Caption:** I was most surprised by the United States and the United Kingdom. In 1952, they already had some of the highest life expectancies in the world (around 69 and 68 years), but by 2007 they had only risen to about 78–79 years. Meanwhile, countries like Japan, Singapore, and Hong Kong made much larger gains and ended up surpassing them.

# Research Discovery & Critical Analysis

## Choose Your Research Question

**Selected Research Question:** *\[A, B, C, D, or E\]*

*\[A\]*

## Research Analysis

### Data Analysis

```{r}
#| label: research-analysis
mid_income_1987 <- gapminder %>%
  filter(year == 1987, gdpPercap >= 3000, gdpPercap <= 12000) %>%
  select(country, continent, gdpPercap_1987 = gdpPercap)

mid_income_status <- gapminder %>%
  filter(year == 2007, country %in% mid_income_1987$country) %>%
  select(country, continent, gdpPercap_2007 = gdpPercap)

mid_income_trap <- left_join(mid_income_1987, mid_income_status, by = c("country", "continent")) %>%
  filter(gdpPercap_2007 < 12000)

print(mid_income_trap)

```

### Visualization

```{r}
#| label: research-plot
#| fig-width: 8
#| fig-height: 6

# Your code here - create one clear plot that supports your findings
ggplot(mid_income_trap, aes(x = gdpPercap_1987, y = gdpPercap_2007, label = country, color = continent)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5, size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Middle-Income Trap (1987–2007)",
       x = "GDP per Capita in 1987",
       y = "GDP per Capita in 2007") +
  theme_minimal()
```

### Interpretation (150-200 words)

This analysis shows that many countries reached middle-income status by 1987 but failed to become high-income by 2007. Latin America is the clearest case, where growth slowed despite earlier progress. This suggests that the middle-income trap is linked to structural issues such as weak institutions, lack of innovation, or dependence on commodities. For example, several Latin American countries remained stuck because they could not upgrade their industries or reduce inequality. While this analysis only looks at GDP per capita and misses social or political factors, it highlights how hard it is to sustain growth once basic industrialization is achieved.
